---
title: "Divvy_Project"
output: html_document
date: "2023-12-01"
---

```{r}
setwd("~/Desktop")
library(data.table)
library(tidyverse)
library(dplyr)
divvy <- fread("Divvy_Trips.csv")
```

Data Cleaning
```{r}
library(janitor)
library(lubridate)
#make everything lowercase and turn spaces in titles into underscores
divvy <- divvy %>% clean_names()
colnames(divvy)

#convert from character to datetime
divvy$start_time <- mdy_hms(divvy$start_time)
divvy$stop_time <- mdy_hms(divvy$stop_time)
```

QUESTION 1
How much of the data does not have an entry for gender? What are the possible reasons for this?
```{r}
unspesified_n <- nrow(divvy %>% filter(gender==""))
unspecified_prop <- unspesified_n/nrow(divvy)
unspecified_prop 
#23.04% of entries do not have gender specified

unspecified <- divvy %>% filter(gender=="") %>% 
  group_by(user_type) %>% 
  summarise(counts =n()) %>%
  mutate(prop = counts/sum(counts))
unspecified
#most unspecified gender entries are for customers. It makes sense that this data is not provided because customers buy passes as needed where as subscribers would be expected to be frequent users with accounts where they enter their personal information
```

QUESTION 2
Plot the number of riders per day stratified by user type. Then by gender.
```{r}
divvy_user <- divvy %>%
  mutate(start_date = as_date(start_time)) %>%
  group_by(start_date, user_type) %>%
  summarize(rides = n()) %>%
  mutate(dayofweek = wday(start_date)) %>%
  group_by(dayofweek, user_type) %>% 
  summarize(ave_ride = mean(rides))

library(ggplot2)

ggplot(divvy_user,
       aes(x=dayofweek, y=ave_ride, color=user_type))+
  geom_line() +
  labs(x = "Day of the Week", y = "Number of Rides", color = "User Type") +
  scale_x_continuous(breaks = 1:7, labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

#why is Dependent so close to 0?
prop_dependent <- nrow(divvy %>% filter(user_type == "Dependent"))/nrow(divvy)
format(prop_dependent*100,  scientific = FALSE)
#less than 1% of rides are by dependents
```

```{r}
#by gender
divvy_gender <- divvy %>%
  filter(gender %in% c("Female", "Male")) %>%
  mutate(start_date = as_date(start_time)) %>%
  group_by(start_date, gender) %>%
  summarize(rides = n()) %>%
  mutate(dayofweek = wday(start_date)) %>%
  group_by(dayofweek, gender) %>% 
  summarize(ave_ride = mean(rides))


ggplot(divvy_gender,
       aes(x=dayofweek, y=ave_ride, color=gender))+
  geom_line() +
  labs(x = "Day of the Week", y = "Number of Rides", color = "Gender") +
  scale_x_continuous(breaks = 1:7, labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```


QUESTION 3
Average number of riders at different times throughout each day. Separate graphs for different genders. 
```{r}
divvy_days <- divvy %>%
  filter(gender %in% c("Female", "Male")) %>%
  mutate(start_date = as_date(start_time),
         s_time = format(start_time, "%H")) %>%
  group_by(start_date, gender, s_time) %>%
  summarize(rides = n()) %>%
  mutate(dayofweek = weekdays(start_date)) %>%
  group_by(dayofweek, s_time, gender) %>%
  summarize(ave_riders = mean(rides)) %>%
  ungroup()

order_of_days <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

#filter for male entries
divvy_days_male <- divvy_days %>% filter(gender=="Male")
#male heat map
ggplot(divvy_days_male,
       aes(x = factor(dayofweek, levels = order_of_days), y=s_time, fill=ave_riders)) + 
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", limits = range(0,max(divvy_days_male$ave_riders))) +
  labs(x = "Day of Week", y = "24-Hour Time", title = "Male Riders per Day of the Week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#filter for female entries
divvy_days_female <- divvy_days %>% filter(gender=="Female")
#female heat map
ggplot(divvy_days_female,
       aes(x = factor(dayofweek, levels = order_of_days), y=s_time, fill=ave_riders)) + 
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(x = "Day of Week", y = "24-Hour Time", title = "Female Riders per Day of the Week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#female heatmap with same range as male
ggplot(divvy_days_female,
       aes(x = factor(dayofweek, levels = order_of_days), y=s_time, fill=ave_riders)) + 
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", limits = range(0,max(divvy_days_male$ave_riders))) +
  labs(x = "Day of Week", y = "24-Hour Time", title = "Female Riders per Day of the Week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

QUESTION 4
During the busiest month, which are the top 10 stations that have more bikes taken than returned. And which more returned than taken.
```{r}
#add variable for start month and variable for stop month
divvy_month <- divvy %>% 
  mutate(start_month = month(start_time), stop_month = month(stop_time))

#check month where most bikes are taken
busy_start <- divvy_month %>% 
  group_by(start_month) %>% 
  summarise(starts_n = n()) %>%
  arrange(-starts_n) %>%
  head(1)
busy_start

#check month where most bikes are returned
busy_stop <- divvy_month %>% 
  group_by(stop_month) %>% 
  summarise(stops_n = n()) %>%
  arrange(-stops_n) %>%
  head(1)
busy_stop

#August (8) is busiest for starting and stopping rides
divvy_month <- divvy_month %>% filter(start_month == 8)

#stations rides start in August
take <- divvy_month %>% 
  group_by(from_station_id, from_station_name) %>% 
  summarise(bikes_taken = n())

#stations rides stop in August
return <- divvy_month %>% 
  group_by(to_station_id, to_station_name) %>% 
  summarise(bikes_returned = n())

#join taken and returned to show number of bikes and taken and returned from each station in August
take_return <- take %>% 
  left_join(return, by= c("from_station_name" = "to_station_name")) %>% 
  #calculate difference between bikes taken and returned
  mutate(diff = bikes_taken-bikes_returned) %>%
  #rename variables
  rename(station_id = from_station_id) %>%
  rename(station_name = from_station_name) %>% 
  #select relevant columns
  select(station_id, station_name, bikes_taken, bikes_returned, diff) 

more_taken <- take_return %>% arrange(-diff) %>% head(10)
more_taken

more_returned <- take_return %>% arrange(diff) %>% head(10)
more_returned
```

QUESTION 5
What proportion of bikes that start being used that year are not utilized in the following years?
```{r}
bike_life <- divvy %>%
  group_by(bike_id) %>% 
  summarize(first_year = min(year(start_time)), 
            last_year = max(year(stop_time))) %>% 
  filter(first_year != 2019)

new_year <- bike_life %>% 
  group_by(first_year) %>%
  summarize(new_bikes = n())
bike_gone <- bike_life %>% 
  filter(first_year==last_year) %>% 
  group_by(first_year) %>%
  summarize(bikes_gone = n()) %>%
  left_join(new_year, by = "first_year") %>%
  mutate(percent_bikes_gone = (bikes_gone/new_bikes)*100)
bike_gone
```
