---
title: "Divvy_Project"
output: html_document
date: "2023-12-01"
---

```{r}
setwd("~/Desktop")
library(data.table)
library(tidyverse)
library(dplyr)
divvy <- fread("Divvy_Trips.csv")
```

Data Cleaning
```{r}
library(janitor)
library(lubridate)
#make everything lowercase and turn spaces in titles into underscores
divvy <- divvy %>% clean_names()
colnames(divvy)

#convert from character to datetime
divvy$start_time <- mdy_hms(divvy$start_time)
divvy$stop_time <- mdy_hms(divvy$stop_time)
```

How much of the data does not have an entry for gender? What are the possible reasons for this?
```{r}
unspesified_n <- nrow(divvy %>% filter(gender==""))
unspecified_prop <- unspesified_n/nrow(divvy)
unspecified_prop 
#23.04% of entries do not have gender specified

unspecified <- divvy %>% filter(gender=="") %>% 
  group_by(user_type) %>% 
  summarise(counts =n()) %>%
  mutate(prop = counts/sum(counts))
unspecified
#most unspecified gender entries are for customers. It makes sense that this data is not provided because customers buy passes as needed where as subscribers would be expected to be frequent users with accounts
```

Number of riders per day stratified by gender. Excluding data missing gender.
```{r}
divvy_week <- divvy %>%
  filter(gender %in% c("Female", "Male")) %>%
  mutate(start_date = as_date(start_time)) %>%
  group_by(start_date, gender) %>%
  summarize(rides = n()) %>%
  mutate(dayofweek = wday(start_date)) %>% 
  group_by(dayofweek, gender) %>% 
  summarize(ave_ride = mean(rides))

library(ggplot2)

ggplot(divvy_week,
       aes(x=dayofweek, y=ave_ride, color=gender))+
  geom_line() +
  labs(x = "Day of the Week", y = "Number of Rides", color = "Gender") +
  scale_x_continuous(breaks = 1:7, labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```


Number of riders at different times throughout each day. Graph for each listed gender

Graph the popularity of Which day of week vs time use heatmap partition by gender
What is the proportion of female to male riders in each month? Rank in descending order. Maybe by starting area? Or by ending area?
  - this information could be valuable for advertising purposes as well as design purposes. Different shapes of seats?
```{r}
divvy_days <- divvy %>% head(1000)
  filter(gender %in% c("Female", "Male")) %>%
  mutate(dayofweek = weekdays(start_time),
         s_time = format(start_time, "%H")) %>%
  group_by(dayofweek, s_time) %>%
  summarize(female = sum(gender=="Female"), 
           male= sum(gender=="Male")) %>%
  ungroup()

ggplot(divvy_days,
       aes(x=dayofweek, y=s_time, fill=female)) + 
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(x = "Day of Week", y = "", title = "Female Riders per Day of the Week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(divvy_days,
       aes(x=dayofweek, y=s_time, fill=male)) + 
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(x = "Day of Week", y = "", title = "Male Riders per Day of the Week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



What are the top 10 most popular start/stop stations?
```{r}
divvy %>% group_by(from_station_name, to_station_name) %>% summarize(riders = n()) %>% arrange (-riders)
```

  
Which station do most rides start from?
  - end of day, need to ensure bikes are brought back to there
  - may also need to pick up bikes from busy stations drop off
```{r}

```

What is the average trip duration per age range (0-10; 10-20; 20-30 etc.)? Create a data visualization stratified by user type.
  - frequency of each age range to get an idea of how many bikes need to be available at a time. Maybe add more bikes to some stations?
  
During the busiest month, which are the top 10 stations that have more bikes taken than returned. And which returned more than taken.
```{r}
divvy_month <- divvy %>% 
  mutate(start_month = month(start_time), stop_month = month(stop_time))

#check month where most bikes are taken
busy_start <- divvy_month %>% 
  group_by(start_month) %>% 
  summarise(starts_n = n()) %>%
  arrange(-starts_n) %>%
  head(1)
busy_start

#check month where most bikes are returned
busy_stop <- divvy_month %>% 
  group_by(stop_month) %>% 
  summarise(stops_n = n()) %>%
  arrange(-stops_n) %>%
  head(1)
busy_stop

#August (8) is busiest for starting and stoping rides
divvy_month <- divvy_month %>% filter(start_month == 8)

take <- divvy_month %>% 
  group_by(from_station_id, from_station_name) %>% 
  summarise(bikes_taken = n())

return <- divvy_month %>% 
  group_by(to_station_id, to_station_name) %>% 
  summarise(bikes_returned = n())

take_return <- take %>% 
  left_join(return, by= c("from_station_name" = "to_station_name")) %>% 
  mutate(diff = bikes_taken-bikes_returned) %>%
  rename(station_id = from_station_id) %>%
  rename(station_name = from_station_name) %>% 
  select(station_id, station_name, bikes_taken, bikes_returned, diff) 

more_taken <- take_return %>% arrange(-diff) %>% head(10)
more_taken

more_returned <- take_return %>% arrange(diff) %>% head(10)
more_returned
```


Of the riders who keep a bike overnight, graph the proportion of riders who keep the bike for <1 day, 1-10 days, 10-20 days, 20-30 days >30 days.
```{r}
divvy_overnight <- divvy %>% 
  select(trip_id, start_time, stop_time, trip_duration) %>%
  mutate(start_day = wday(start_time), stop_day = wday(stop_time)) %>%
  filter(start_day != stop_day) %>% 
  #calculate number of days
  mutate(trip_days = format(trip_duration/(60*60*24), scientific = FALSE))
#create day categories
#graph
```


Maybe which bikes are out of commission?
How long a bike is kept?

How many bikes from each year are not utilized in the following years?
```{r}
divvy %>%
  group_by(bike_id) %>% 
  summarize(first_year = min(year(start_time)), 
            last_year = max(year(stop_time))) %>% 
  mutate(lifetime = last_year - first_year) %>% 
  filter(arrange(-lifetime))

new_per_year <- 
  divvy %>%
  mutate(f_year = year(start_time), l_year = year(stop_time)) %>%
  filter(f_year != 2019) %>%
  group_by(f_year) %>%
  summarize(numbers = n())

divvy %>%
  mutate(f_year = year(start_time), l_year = year(stop_time)) %>%
  mutate(lifetime = l_year - f_year) %>%
  filter(lifetime ==0) %>%
  group_by(f_year) %>%
  summarize(numbers = n())

(first_year = min(year(start_time)), 
            last_year = max(year(stop_time))) %>% 
  filter(arrange(-lifetime)
```


